---
title: "A Multi-Metric Analysis of Ecosystem Recovery"
author: "Gabriela Albuquerque Galvão"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)

library(tidyverse)
library(ggplot2)
library(vegan)
library(ade4)
library(FD)
library(knitr)
library(kableExtra)
```
### Packages

- tidyverse  
- ggplot2  
- vegan  
- ade4  
- FD  
- knitr  
- kableExtra  


---

## Executive summary

> **The core question:** Can a single metric tell the full story of a complex system, or does it hide what matters most?

This project analyzes biodiversity data collected from areas impacted by the São Francisco River transposition project (Caatinga biome, Brazil). Using a **multi-metric analytical framework**, I demonstrate that surface-level indicators can be misleading, that is: a finding directly applicable to any domain where data-driven decisions depend on choosing the right KPI.

**Stack:** R · tidyverse · ggplot2 · vegan · FD · multivariate statistics · mixed models

**Key result:** Taxonomic richness showed *no significant difference* between groups. Functional diversity revealed *highly significant differences* (p < 0.01). Same data. Different question. Different answer.

---

## 1. The Problem

Environmental restoration projects, like business interventions, require monitoring tools to evaluate their effectiveness. But which metric do you trust?

This dataset covers **four distinct landscape conditions**, analogous to experimental groups in an A/B test:

| Group | Description |
|---|---|
| **Control** | Non-impacted reference areas |
| **Degraded** | Impacted areas of São Francisco river transposition |
| **Restoration 1** | Direct seeding technique |
| **Restoration 2** | Combined restoration techniques |

The goal: determine whether restoration is working, but also and *how* to measure "working" in an ecology context.

---

## 2. Data

```{r echo=TRUE}
dataset_cmm <- read.csv("comunidade.csv",
                       sep = ";",
                       row.names = 1)
kable(head(dataset_cmm[, 1:10]), caption = "Community matrix (first 10 columns shown)") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"))

traits <- read.csv("traits.csv",
                   sep = ";",
                   row.names = 1,
                   fileEncoding = "Latin1")
kable(head(traits), caption = "Species trait matrix") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"))

# Species abundance matrix
cmm_table <- dataset_cmm %>%
  dplyr::select(21:126)

# Treatment groups
treatments <- dataset_cmm %>%
  dplyr::pull(Grupo)
```

**Dataset dimensions:** `r nrow(cmm_table)` sampling units × `r ncol(cmm_table)` species, plus `r ncol(traits)` functional trait variables per species.

---

## 3. Data Preparation & Feature Engineering

Standardization is essential before multivariate analysis. Here I apply the **Hellinger transformation**, it down-weights highly abundant species and normalizes sampling effort, making samples comparable regardless of total seed counts.

> *Business analogy: this may be equivalent to normalizing revenue by market size before comparing regions.*

```{r echo=TRUE}
# Hellinger transformation
cmm_hell <- vegan::decostand(cmm_table, method = "hellinger")

# Feature 1: taxonomic richness (species count per sample)
richness <- rowSums(cmm_table > 0)

dados <- data.frame(
  richness = richness,
  group = treatments
) %>%
  mutate(
    group = case_when(
      group == "N"   ~ "Degraded",
      group == "P"   ~ "Control",
      group == "RE1" ~ "Restoration 1",
      group == "RE2" ~ "Restoration 2"
    ),
    group = factor(group, levels = c("Control", "Degraded", "Restoration 1", "Restoration 2"))
  )
```

---

## 4. Metric 1: Taxonomic richness (the obvious KPI)

The first and most intuitive metric: how many species are present in each condition?

```{r echo=TRUE}
ggplot(dados, aes(x = group, y = richness, fill = group)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA, color = "black") +
  geom_jitter(width = 0.12, alpha = 0.5, size = 2) +
  stat_summary(fun = mean, geom = "point", size = 3, color = "black") +
  scale_fill_manual(values = c(
    "Control"       = "#33a02c",
    "Degraded"      = "#ca0020",
    "Restoration 1" = "#1f78b4",
    "Restoration 2" = "#6A5ACD"
  )) +
  theme_minimal(base_size = 12) +
  labs(
    title = "Species Richness by Landscape Condition",
    subtitle = "Points = individual samples | Black dot = group mean",
    x = NULL, y = "Number of Species"
  ) +
  theme(legend.position = "none", plot.title = element_text(face = "bold"),
        axis.text.x = element_text(angle = 20, hjust = 1))
```

```{r echo=TRUE}
anova_richness <- aov(richness ~ group, data = dados)
shapiro.test(residuals(anova_richness))
bartlett.test(richness ~ group, data = dados)
summary(anova_richness)
```

**Result:** ANOVA assumptions met. No statistically significant difference detected (p > 0.05)

**Interpretation:** If this were the only metric in a dashboard, the conclusion would be: *"Restoration has no effect."*. But that conclusion would be wrong.

---

## 5. Metric 2 — Functional diversity (The better KPI)

Species richness counts *how many*. Functional diversity measures *what kind*, capturing the range of ecological roles present in each community.

**FDis (Functional Dispersion):** measures how spread species are in trait space. Higher FDis = more functionally diverse community = more resistent ecosystem.

> *Business analogy: the difference between counting the number of employees vs. measuring the diversity of skills on a team.*

```{r echo=TRUE}
traits_num <- traits %>%
  mutate(h_max = as.numeric(as.character(h_max))) %>%
  select(h_max) %>%
  drop_na()

common_species <- intersect(rownames(traits_num), colnames(cmm_table))
traits_num    <- traits_num[common_species, , drop = FALSE]
cmm_aligned   <- cmm_table[, common_species]
cmm_hell      <- vegan::decostand(cmm_aligned, method = "hellinger")
traits_num    <- traits_num[colnames(cmm_hell), , drop = FALSE]

fdis      <- FD::dbFD(traits_num, cmm_hell)$FDis
dados$fdis <- fdis

ggplot(dados, aes(x = group, y = fdis, fill = group)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA, color = "black") +
  geom_jitter(width = 0.12, alpha = 0.5, size = 2) +
  stat_summary(fun = mean, geom = "point", size = 3, color = "black") +
  scale_fill_manual(values = c(
    "Control"       = "#33a02c",
    "Degraded"      = "#ca0020",
    "Restoration 1" = "#1f78b4",
    "Restoration 2" = "#6A5ACD"
  )) +
  theme_minimal(base_size = 12) +
  labs(
    title = "Functional diversity (FDis) by landscape condition",
    subtitle = "FDis captures the 'profile diversity' of the community: not just its size",
    x = NULL, y = "Functional dispersion (FDis)"
  ) +
  theme(legend.position = "none", plot.title = element_text(face = "bold"),
        axis.text.x = element_text(angle = 20, hjust = 1))

anova_fdis <- aov(fdis ~ group, data = dados)
shapiro.test(residuals(anova_fdis))
bartlett.test(richness ~ group, data = dados)
summary(anova_fdis)
```

**Result:** Significant differences detected (p < 0.01).

**Interpretation:** Restoration *is* working, but only a deeper metric can see it.

---

## 6. Metric 3 — Community Composition (Structural Analysis)

PCoA (Principal Coordinates Analysis) is a dimensionality reduction technique, analogous to PCA, that maps community similarity in 2D space. Points closer together = more similar composition.

```{r echo=TRUE}
dist_matrix <- vegdist(cmm_hell, method = "bray")
pcoa        <- cmdscale(dist_matrix, k = 2)

df_pcoa <- data.frame(
  Axis1 = pcoa[, 1],
  Axis2 = pcoa[, 2],
  group = dados$group
)

ggplot(df_pcoa, aes(Axis1, Axis2, color = group)) +
  geom_point(size = 3.5, alpha = 0.85) +
  stat_ellipse(aes(fill = group), geom = "polygon", alpha = 0.08, linetype = 2) +
  scale_color_manual(values = c(
    "Control"       = "#33a02c",
    "Degraded"      = "#ca0020",
    "Restoration 1" = "#1f78b4",
    "Restoration 2" = "#6A5ACD"
  )) +
  scale_fill_manual(values = c(
    "Control"       = "#33a02c",
    "Degraded"      = "#ca0020",
    "Restoration 1" = "#1f78b4",
    "Restoration 2" = "#6A5ACD"
  )) +
  theme_minimal(base_size = 12) +
  labs(
    title = "Community Composition Space (PCoA)",
    subtitle = "Separation between groups = structural differences invisible to richness",
    x = "Principal Coordinate 1", y = "Principal Coordinate 2",
    color = NULL, fill = NULL
  ) +
  theme(plot.title = element_text(face = "bold"))
```

**Interpretation:** Even when counts look similar, communities can occupy completely different positions in composition space — different species, different functional roles, different ecological dynamics.

---

## 7. Insights & Takeaways

### Finding 1 - The wrong KPI gives the wrong answer

| Metric | Result | Conclusion |
|---|---|---|
| Species Richness | p > 0.05, ns | "No effect of restoration" ❌ |
| Functional Diversity | p < 0.01 ** | "Restoration works" ✅ |
| Composition (PCoA) | Visual separation | "Groups are structurally different" ✅ |

> **Data lesson:** A dashboard showing "no change" in a top-line metric may be hiding significant structural shifts. Always interrogate *what* you're measuring, not just *whether* it changed.


### Finding 2 - Structure ≠ Volume

Two groups can have the same number of items (species, customers, transactions) while being entirely different in *composition*. Multivariate methods reveal this. Aggregate counts don't.

> **Data lesson:** Segment-level structure analysis (clustering, PCA, PCoA) catches what KPI dashboards miss.

---

## Skills Demonstrated  
*(AI generated based on the script)*

| Skill                     | Applied as                                                                 |
|--------------------------|---------------------------------------------------------------------------|
| Data wrangling           | tidyverse pipelines, matrix alignment, missing data handling              |
| Feature engineering      | Hellinger transformation, functional metrics (FDis)                       |
| Statistical testing      | ANOVA with assumption checking (Shapiro, Bartlett)                        |
| Dimensionality reduction | PCoA / cmdscale on Bray-Curtis dissimilarity matrix                       |
| Data visualization       | ggplot2 with layered geometries, ellipses, color encoding                 |
| Critical interpretation  | Multi-metric framework, translating results to decisions                  |
---

*Gabriela Albuquerque Galvão · ggalvao96@gmail.com · Recife, PE*
